Simulating the Linux page cache in distributed applications with the WRENCH library

### 2020-12-09 paper_io_sim.tex
fixes to intro and related work

        The Linux Page Cache, introduced in kernel version 2.4, plays an important role in
        the reduction of filesystem data transfer costs~\cite{pagecache}. With the page cache, previously
        read data can be reread directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance benefits that can be obtained from using the page
        cache are determined by many factors including the total amount of memory,
        the amount of data being written (i.e. dirty data), and the percentage of available memory available for
        written data. All these factors are important when determining the impact of data transfers on
        application performance, particularly in data intensive applications.

        Characterizing the performance of applications directly on computing infrastructure
        has various shortcomings, which include, limited and shared resources, operational
        costs, unstable environments, and the difficulties in producing reproducible results.
        Simulations address these concerns by providing abstractions to the performance of computer
        hardware, such as CPU, network and storage. As a result, simulations provide and
        cost-effective, fast, easy and reproducible  way evaluate application performance
        on different infrastructures without having to execute the code.

        Page cache simulation is crucial when simulating data intensive applications,
        which would rely on cache hits and writing to memory for additional performance.
        While existing simulators can adequately characterize many different computing environments,
        they lack much of the ability to adequately describe the page cache. Some
        simulators try to simulator I/O with page cache, but fail to capture the key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Other simulators, such as~\cite{xu2018}, capture the important feature of the
        page cache, but are domain-specific.

@@

        In this paper, we present WRENCH-cache, a simulation model of I/O with page
        cache incorporated into WRENCH, a workflow management system simulator based of
        the popular SimGrid framework. The goal of WRENCH-cache is to provide users
        with an accurate representation of a workflow execution, when file system page
        cache is used. WRENCH-cache is able to simulate both writeback and writethrough
        cache which would normally be available to local and NFS-based filesystems.

@@

            Linux implements a disk cache called \textit{page cache}. 
            It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to avoid disk I/O by keeping data in memory 
            for later accesses, minimize I/O time since disk accesses are considerably 
            slower than memory accesses.
            When the kernel starts a read operation, it checks if required data is in page cache.
            If yes, data is read from memory. This is called a \textit{cache hit}. 
            Otherwise, data is read from disk and placed in page cache.
            For writing, page cache is implemented with three different strategies. 
            The first is \textit{direct~I/O} or \textit{no-write}, which means the page cache 
            is not involved in write operations. 
            The second is \textit{write-through}, in which data goes \textit{through} 
            page cache to disk. In this case, data is written to both page cache and disk. 
            In the third strategy, called \textit{writeback}, data are written directly 
            to disk as \textit{dirty} pages and these pages will be periodically flushed 
            to disk later by a \textit{writeback} process and then marked as not dirty. 
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes and perform a bulk write later.
            \cite{linuxdev3rd2010}.

            In writeback, one of the keys is the cache eviction mechanism, which is
            triggered when available memory is insufficient to removes pages 
            from page cache. 
            Because dirty pages are data written to page cache but yet to be 
            written to disk, dirty pages must be kept in page cache and only clean pages 
            are available for eviction.
            If the clean pages are inadequate, dirty pages are flushed to disk, 
            marked as clean and available for eviction. 
            There is another type of flushing mechanism, which is 
            \textit{periodical flushing}, called periodically in predefined intervals. 
            Periodical flushing only flushes expired dirty pages, which remain dirty in 
            page cache longer than an expiration time configured in the kernel. 
            Different cache eviction algorithms have also been developed and proposed
            \cite{owda2014comparison}.
            In Linux kernel, a \textit{two lists strategy (LRU/2)} is implemented based on 
            the least-recently-used (LRU) scheme, with an \textit{active list} and 
            an \textit{inactive list}. 
            If accessed pages are not in the page cache, they are added to the inactive list. 
            If pages are accessed when residing in the inactive list, they will be moved from 
            inactive list to active list. 
            The lists are also kept balanced by moving pages from the active list
            to the inactive list when the active list grows too large.
            Thus, the active list only contains pages which are accessed more than once 
            and not evictable, while the inactive list includes pages accessed once only, 
            or pages accessed more than once but moved from the active list. 
            The lists work in LRU manner, pages are added to the tail and removed 
            from the top.

@@

            Many simulation frameworks and applications have been developed 
            in recent years to support studies in grids, cloud, HPC, peer-to-peer 
            networks, volunteer computers.                     
            These simulators are built with models of resources such as CPU, network, 
            storage with different approaches and levels of details in different simulators. 
            The detail level of the model causes a conflict between accuracy and scalability, 
            and these simulators usually trade of one for the other 
            \cite{casanova2014simgrid}. 
            For example, some simulators are built with packet-level network models 
            such as \textit{CloudSim} and \textit{iCanCloud}, or storage model that 
            simulates the operation of storage hardware in \textit{DiskSim} 
            \cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but at the same time lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, which allows the use of the framework 
            to simulate applications in multiple domains \cite{casanova2014simgrid}.

            Despite the fact that many sophisticated models for resources have been 
            integrated in simulators, and the page cache shows apparent impacts on I/O
            performance, the simulation model of page cache is still missing in most 
            of simulators.
            For example, SimGrid only simulates I/O operations with storage bandwidths 
            and capacity. 
            SIMCAN is a simulation framework that models page cache by storing data 
            accessed on disk in a block cache \cite{nunez2012simcan}. 
            Page cache is also modeled in iCanCloud simulator with a component in charge 
            of managing memory accesses and cached data \cite{nunez2012icancloud}, 
            but one of the issues is scalability due to its sophisticated model.
            Ultimately, non of these simulators provide any cache write strategy, 
            cache eviction mechanism with dirty/clean data and the page cache LRU lists. 
            Cache replacement policies is applied in \cite{xu2018saving} to simulate 
            in-memory caching, but for the study in energy consumption of multi-tier 
            heterogenous network.

            In this study, we chose SimGrid to implement our model of I/O with page cache 
            due to the balancing between accuracy and scalibility, the versatility allowing it 
            to be adopted in different domains. 
            In addition, it is an actively maintained open-scource framework with 
            modular design, which enables the framework to be extended with new models. 
            The main issue users have to face when using SimGrid is everything has to be 
            implemented from scratch due to its low-level API. 
            To ease this challenge, WRENCH, another open-source library, is built on top 
            of SimGrid to abstract the low-level logic and expose a higher abstraction 
            level API \cite{casanova2018wrench}. 
            We integrate our model in WRENCH and use the SimGrid API to interact with 
            simulated storage resource.
            This approach allows us to leverage existing features in WRENCH and SimGrid.
            Firstly, we can take advantage of the process control, synchronization of 
            resource consumption by concurrent processes provided by the simulation core 
            in SimGrid. 
            Secondly, we can reuse the existing SimGrid storage model with capacity 
            and bandwidth properties, which are similar to the characteristics of memory.
            And finally, WRENCH provides a high level abstraction, which help us drop 
            the burden of implementation.

    \section{Method}
    \label{method}
    In this section, we present our page cache simulation model for file I/O,
    flusing, and eviction. We implement this model in two simulators:
    (1) a standalone Python prototype, and (2) WRENCH-cache, an extension of the WRENCH library.
    We also describe experimental scenarios to evaluate our simulators in various configurations.

    We separate our simulation model in two components, the IO
    Controller and the Memory Manager, which together simulate 
    file reads and writes in Linux (Figure~\ref{fig:interaction}). 
    To read or write a file chunk, a simulated application sends a
    request to the IO Controller. The IO Controller interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write to cache. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction from cache, and reads or writes to disk when necessary.
    \begin{figure}
           \centering
           \includegraphics[width=0.85\columnwidth]{figures/interaction.pdf}
           \caption{Overview of the simulated writeback page cache \tristan{does it simulate only writeback? See comment in NFS section.}.
           Applications send file read or write requests to the
           IO Controller that orchestrates flushing, eviction, cache
           and disk accesses with the Memory Manager. Concurrent accesses to storage
           devices (memory and disk) are simulated using existing models.}
           \label{fig:interaction}
    \end{figure}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cache I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate disk and
    memory, characterized by its storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses and concurrent disk accesses.

@@

    We model page cache LRU lists as 
    two lists of data blocks, an active list and an inactive list, both ordered by 
    last access time (earliest first, see Figure~\ref{fig:lrulist}).
    As in the kernel, our simulator limits the size of the active list to
    twice the size of the inactive list, by moving least recently 
    used data blocks from the active list to the inactive list~\cite{gorman2004understanding, linuxdev3rd2010}.
    % Modeling page cache LRU lists as lists of data blocks reduces the
    % overhead of the simulator, as data blocks are obviously less
    % numerous than file pages. At the same time, it achieves the accuracy in 
    % I/O time simulation that can be seen in the later section.
    % From Tristan: I don't think the sentences above are necessary,
    % they are quite repetitive.

@@

    \subsubsection{Reads and writes}

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are 
    accessed in a round-robin fashion rather than fully randomly. 
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (data reads occur from left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created 
    and appended to the inactive list. 
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block, 
    and append it to the active list. 
    If the blocks are dirty, we move them independently to the active list, to preserve their entry time. 
    Because the chunk and block sizes may be different, there are situations
    where a block is not entirely read. 
    In this case, the block is split in two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data 
           is read first, followed by data from the inactive list, and finally data from the active list. \tristan{Colors 
           shouldn't match that in Fig 1 as they dont represent the same thing.}}
           \label{fig:read_order}
    \end{figure}

    For file writes, we assume that all data to be written is 
    uncached. Thus, each time a chunk is written, we create a block of dirty data 
    and append it to the inactive list.

@@

    The main simulated thread in the Memory Manager can flush or evict data from the
    memory cache. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk, having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. Flushing time
    to disk is simulated by the storage model.

    The cache eviction simulation also runs in
    the main thread. It frees up the page cache by traversing and deleting 
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted 
    from the inactive list until the evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The cache eviction simulation does not contribute to the simulated time 
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if (1) it is dirty, and (2)
    the duration since its entry time is longer than a
    predefined expiration time. 
    Periodical flushing is simulated as an infinite loop in which 
    the Memory Manager searches and flushes these expired dirty blocks to disk (Algorithm~\ref{alg:pdflush}). 
    % From Tristan: the algorithm is quite straightforward, I don't think this is 
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two 
    % page cache LRU lists (line 8), simulates writes of data of these blocks 
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put 
    % to sleep for the remaining time (lines 11-13). 
    % Then based on the state of the host at current simulated time, 
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This is taken into account by the 
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation of IO Controller}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assumed to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn)) \Comment{To be read from disk} 
               \State cache\_read = cs - disk\_read \Comment{To be read from cache} 
            \State mm.flush(cs + disk\_read - mm.free\_mem - mm.evictable, fn) 
            \State mm.evict(cs + disk\_read - mm.free\_mem, fn) 
            \If {disk\_read $>$ 0}  \Comment{Read uncached data}    
                \State sm.read(disk\_read)  
                \State mm.add\_to\_cache(disk\_read, fn)     
            \EndIf
            \If {cache\_read $>$ 0} \Comment{Read cached}
                \State mm.cache\_read(cache\_read)  
            \EndIf
        \end{algorithmic}
    \end{algorithm}
    As mentioned previously, our model reads and writes file chunks in a
    round-robin fashion. To read a file chunk, simulated applications send
    chunk read requests to the IO Controller which processes them using
    Algorithm~\ref{alg:read}. First, we calculate the amount of uncached
    data that needs to be read from disk, and the remaining amount is read
    from cache (line 7-8). The amount of memory required to read the chunk
    is calculated, corresponding to a copy of the chunk in anonymous memory
    and a copy of the chunk in cache \tristan{Create a variable
    "required\_mem" in the algorithm to clarify}.
    \tristan{add a call to mm in the algorithm to increment anonymous memory. Mention in the experiments that anonymous memory is flushed after each task.}
    If there is not enough available memory, the Memory Manager is called
    to flush dirty data (line 9). If necessary, flushing is complemented by
    eviction (line 10). Note that, when called with negative arguments, functions
    \texttt{flush} and \texttt{evict} simply return and do not do anything. Then, if the block requires
    uncached data, a disk read is simulated with the disk model (line 12)
    \tristan{this is not consistent with the new Figure 1 where disk reads
    are done by the mm}, and the data read is added to cache (line
    13).
    Finally, if cached data needs to be read, the Memory Manager is called
    to simulate a cache read  
    and update the corresponding data blocks accordingly (line 16).

    \begin{algorithm}\caption{File chunk write simulation of IO Controller}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write(fn, mem\_amt) 
            \EndIf
            \If {mem\_amt $<$ cs}  \Comment{Write with disk bandwidth}
                \State mm.flush(cs - mem\_amt)  
                \State mm.evict(cs - mem\_amt  - mm.free\_mem) 
                \State mm.add\_to\_cache(fn, min(cs - mem\_amt, mm.free\_mem)) \tristan{why is data written to cache with "addtocache" here and with "write" at line 10?}
                \State sm.write(cs - mem\_amt) \tristan{According to Fig 1 the IO Controller is not supposed to interact directly with sm}
            \EndIf

>>>>

        The Linux Page Cache plays an important role in
        the reduction of filesystem data transfer costs. With the page cache, previously
        read data can be reread directly from memory and written data can be written to
        memory before being asynchronously flushed to disk, resulting in improved I/O performance
        on slower storage devices. The performance benefits that can be obtained from using the page
        cache are determined by many factors including the total amount of memory,
        the amount of data being written (i.e. dirty data), and the percentage of available memory available for
        written data. All these factors are important when determining the impact of data transfers on
        application performance, particularly in data intensive applications.

        Characterizing the performance of applications directly on computing infrastructure
        has various shortcomings, which include, limited and shared resources, operational
        costs, unstable environments, and the difficulties in producing reproducible results.
        Simulations address these concerns by providing abstractions for the performance of computer
        hardware, such as CPU, network and storage. Consequently, simulations provide and
        cost-effective, fast, easy and reproducible  way evaluate application performance
        on different infrastructures without having to execute the code.

        The amount of data-intensive applications has been on the rise as a result of
        open-data and data sharing initiatives. Due to the sheer size of the data being
        processed, these applications must be executed on large scale infrastructure
        such as High Performance Computing (HPC) clusters or the cloud. Determining
        which type of infrastructure is the most cost-effective for the particular
        application can be tricky without the help of a simulator.

        Page cache simulation is crucial when simulating data intensive applications,
        which rely on cache hits and writing to memory for additional performance.
        Furthermore, simulating the page cache enables us to study important application-I/O
        bottlenecks which can arise over the course of execution, such as page cache
        starvation~\cite{zhuang2017}.
        While existing simulators can adequately characterize many different computing environments,
        they lack much of the ability to adequately describe the page cache. Some
        simulators try to simulator I/O with page cache, but fail to capture the key features such
        as dirty data and cache eviction policies~\cite{nunez2012simcan,nunez2012icancloud}. 
        Other simulators, such as~\cite{xu2018}, capture the important feature of the
        page cache, but are domain-specific.

@@

        In this paper, we present WRENCH-Cache, a simulation model of I/O with page
        cache incorporated into WRENCH~\cite{casanova2018wrench}, a workflow management system simulator based on
        the popular SimGrid framework. The goal of WRENCH-Cache is to provide users
        with an accurate representation of a workflow execution, when file system page
        cache is used, while providing a good balance between accuracy and scalability.
        WRENCH-Cache is able to simulate both writeback and writethrough
        cache which would normally be available to local and NFS-based filesystems.
        WRENCH-Cache can provide accurate simulations for large-scale parallel
        applications using the page cache.

@@

        To improve performance of disk accesses, the page cache was introduced to the Linux kernel. %It contains pages referring to physical pages on disk \cite{linuxdev3rd2010}.
            The goal of the page cache is to try ensure that most I/O operations occur entirely
            within memory space rather than synchronously to and from the storage device.
            All disk-backed data located in the page cache are partitioned into many small pages.
            When a read operation is initiated, the kernel checks whether the required pages are in memory.
            If it is the case, data is read entirely from memory (i.e. \textit{cache hit}). Otherwise,
            the necessary pages must be loaded to memory from the disk (i.e. \textit{cache miss})
            Conversely, for writing, one of three strategies can be applied: direct~I/O (or, ``no-write''),
            writethrough and writeback.
            If direct~I/O is enabled for a given file system, all I/O operations bypass the
            page cache for writes, writing data directly to disk. In contrast, write-through enables the page cache
            for file system writes, but the data is written synchronously from memory to disk. Writeback extends write-through
            by enabling asynchronous writes to disk, meaning that a write call will return once
            the data has been written to memory.
            The writeback strategy is considered to outperform write-through as well as 
            direct I/O as it delays disk writes to perform a bulk write at a later time.
            \cite{linuxdev3rd2010}.

            Cache eviction and flushing strategies are integral to proper page cache functioning.
            Whenever space in memory becomes limited, either as a result of application memory
            or page cache use, page cache data may be evicted. Only data that 
            has been persisted to storage (clean pages) can be flagged for eviction and removed from
            memory. Written data that has not yet been persisted to disk (dirty data) must first
            be copied (flushed) to storage prior to eviction. When sufficient memory is
            being occupied, the flushing process is synchronous. However, even when 
            there is sufficient available memory, written data will be flushed to disk
            at a predefined interval through a process known as \textit{periodical flushing}.
            Periodical flushing only flushes expired dirty pages, which remain dirty in 
            page cache longer than an expiration time configured in the kernel. 
            Different cache eviction algorithms have also been developed and proposed
            \cite{owda2014comparison}.
            The Linux kernel uses a two-list strategy to flag pages for eviction.
            The two-list strategy is based on a least recently used (LRU) policy
            and uses an active and inactive list in its implementation.
            If accessed pages are not in the page cache, they are added to the inactive list. 
            Should pages located on the inactive list be accessed, they will be moved from 
            the inactive to the active list. 
            The lists are also kept balanced by moving pages from the active list
            to the inactive list when the active list grows too large.
            Thus, the active list only contains pages which are accessed more than once 
            and not evictable, while the inactive list includes pages accessed once only, 
            or pages that have be accessed more than once but moved from the active list. 
            Both lists operate using LRU eviction policies meaning the data that has
            not be accessed recently will be moved first.

@@

            Many simulation frameworks exist to describe the behaviours of applications
            running on the various infrastructures available to users, such as personal computers,
            clouds and clusters.
            These simulators are built with resource models such CPU, network and 
            storage, using various different approaches to describe the hardware.
            For example, there is \textit{CloudSim} and \textit{iCanCloud} for packet-level network models 
            and \textit{DiskSim} the simulates storage hardware functioning.

            Simulators must choose between scalability and accuracy when describing the
            hardware.\cite{casanova2014simgrid}. 
            Fine-grained models allow simulators to achieve high accuracy, 
            but lead to scalability issues due to long simulation time 
            when simulated platforms and applications grow. 
            SimGrid is an open source simulation platform that has been developed 
            not only to balance the trade-off between accuracy and scalability, 
            but also to achieve versatility, allowing the use of the framework 
            to simulate applications in multiple domains. WRENCH, a workflow management
            system simulator based on the SimGrid framework, provides a high-level API
            for simulating workflows, thus facilitating the integration of new models.

            Despite the fact that many sophisticated models for resources have been 
            integrated in simulators, and the page cache shows apparent impacts on I/O
            performance, the simulation model of page cache is still missing in most 
            of simulators.
            For example, SimGrid only simulates I/O operations with storage bandwidths 
            and capacity. 
            SIMCAN is a simulation framework that models page cache by storing data 
            accessed on disk in a block cache \cite{nunez2012simcan}. 
            Page cache is also modeled in the iCanCloud simulator through a component which
            manages memory accesses and cached data \cite{nunez2012icancloud}. However,
            the scalability of the iCanCloud simulator is limited by its sophisticated model.
            Ultimately, none of these simulators provide any writeback cache simulator nor
            cache eviction policys through LRU lists.
            Although cache replacement policies are applied in \cite{xu2018saving} to simulate 
            in-memory caching, their simulator is specific to energy consumption of multi-tier 
            heterogenous networks.

            In this study, we use WRENCH 
            to implement our page cache model. We chose WRENCH due to accurate yet scalable implementation, resulting
            from its SimGrid backend, as well as its high-level API.
            We have integrated our model in WRENCH and use the SimGrid API to interact with 
            simulated storage resources.
            This approach allows us to leverage existing features in WRENCH and SimGrid such as process control and synchronization of 
            resource consumption.
            We also leverage the existing SimGrid storage model with capacity 
            and bandwidth properties, which are similar to the characteristics of memory.

    \section{Method}
    \label{method}
    In this section, we present our page cache simulation model for file I/O,
    flusing, and eviction. We detail the design and implementation of two simulators:
    (1) a standalone Python prototype to evaluate the simulation model independently of any other 
    tool, and (2) WRENCH-cache, an extension of WRENCH with our page cache model.
    We also describe experimental scenarios to evaluate our simulators in various configurations.

    We separate our simulation model in two components, the IO
    Controller and the Memory Manager, which together simulate 
    file reads and writes in Linux (Figure~\ref{fig:interaction}). 
    To read or write a file chunk, a simulated application sends a
    request to the IOController. The IOController interacts as needed with
    the Memory Manager to free memory through flushing or eviction,
    and to read or write to cache. The Memory Manager
    implements these operations, simulates periodical flushing
    and eviction from cache, and reads or writes to disk when necessary.
    \begin{figure}
           \centering
           \includegraphics[width=0.85\columnwidth]{figures/interaction.pdf}
           \caption{Overview of the simulated writeback page cache.
           Applications send file read or write requests to the
           IOController that orchestrates flushing, eviction, cache
           and disk accesses with the MemoryManager. Storage
           devices (memory and disk) simulate file transfers with
           bandwidth sharing.}
           \label{fig:interaction}
    \end{figure}

@@

    The Memory Manager simulates two parallel threads: the main one
    implements flushing, eviction, and cache I/Os synchronously, whereas
    the second one, operating in the background, periodically searches for
    expired dirty data in page cache LRU lists and flushes them to disk. We
    use existing storage simulation models~\cite{lebre2015} to simulate
    memory, characterized by its storage capacity, read and write
    bandwidths, and latency. Reusing existing storage models allows us to
    simulate bandwidth sharing between concurrent memory accesses.

@@

    We model page cache LRU lists as 
    two lists of data blocks, an active list and an inactive list, both ordered by 
    last access time (earliest first, Figure~\ref{fig:lrulist}).
    As in the kernel, our simulator limits the size of the active list to
    twice the size of the inactive list, by moving least recently 
    used data blocks from the active list to the inactive list~\cite{gorman2004understanding, linuxdev3rd2010}.
    % Modeling page cache LRU lists as lists of data blocks reduces the
    % overhead of the simulator, as data blocks are obviously less
    % numerous than file pages. At the same time, it achieves the accuracy in 
    % I/O time simulation that can be seen in the later section.
    % From Tristan: I don't think the sentences above are necessary,
    % they are quite repetitive.

@@

    \subsubsection{Reads and Writes}

    Our simulation model supports chunk-by-chunk file accesses
    with a user-defined chunk size. However, for simplicity, we assume that file pages are 
    accessed in a round-robin fashion rather than fully randomly. 
    Therefore, when a file is read, cached data is read only after all uncached data was read, and data from the inactive list is read
    before data from the active list
    (left to right in Figure~\ref{fig:read_order}).
    When a chunk of \emph{uncached} data is read, a new clean block is created 
    and put at the top of the inactive list. 
    When a chunk of \emph{cached} data is read, one or more existing data blocks in the LRU lists are accessed.
    If these blocks are clean, we merge them together, update the access time and size of the resulting block, 
    and put it in the active list. 
    If the blocks are dirty, we move them to the active list to preserve their entry time. 
    Because the chunk size and block size can be different, there are situations
    where a block is not entirely read. 
    In this case, the block is split in two smaller blocks and one of them is re-accessed.
    % From Tristan: not sure where to put this nor if it's necessary:
    % , in which chunks are read/written
    % until file is entirely read/written.
    \begin{figure}
           \centering
           \includegraphics[width=\columnwidth]{figures/read_order.pdf}
           \caption{File data read order. Data is read from left to right: uncached data 
           is read first, followed by data from the inactive list, and finally data from the active list.}
           \label{fig:read_order}
    \end{figure}

    For file writes, we assume that all data to be written is 
    uncached. Thus, each time a chunk is written, we create a block of dirty data 
    and add it to the inactive list.

@@

    The main simulated thread in the Memory Manager can flush the
    cache to write dirty data to disk. The data flushing simulation
    function takes the amount of data to flush as parameter. While
    the amount of data to flush is not reached and there are dirty
    blocks remaining in cache, this function traverses the sorted
    inactive list, then the sorted active list, and writes the
    least recently used dirty block to disk having set its dirty
    flag to 0. In case the amount of data to flush requires that a
    block be partially flushed, the block is split in two blocks,
    one that is flushed and one that remains dirty. Flushing time
    to disk is simulated with the storage model.

    Similarly to data flushing, our cache eviction simulation runs in
    the main thread. It frees up the page cache by traversing and deleting 
    least recently used clean data blocks in the inactive list.
    The amount of data to evict is passed as a parameter and data blocks are deleted 
    from the inactive list until total evicted data reaches the required amount,
    or until there is no clean block left in the list.
    If the last evicted block does not have to be entirely evicted, the block is split in two blocks,
    and only one of them is evicted.
    The cache eviction simulation does not contribute to the simulated time 
    since cache eviction time is negligible in real systems \tristan{a reference would be welcome}.

@@

    Periodical flushing is simulated in the Memory Manager
    background thread. As in the Linux kernel, a dirty block
    in our model is considered expired if (1) it is dirty, and (2)
    the duration since its entry time is longer than a
    predefined expiration time. 
    Periodical flushing is simulated as an infinite loop in which 
    the Memory Manager searches and flushes these expired dirty blocks to disk (Algorithm~\ref{alg:pdflush}). 
    % From Tristan: the algorithm is quite straightforward, I don't think this is 
    % necessary
    % In each repetition, Memory Manager finds expired dirty blocks in two 
    % page cache LRU lists (line 8), simulates writes of data of these blocks 
    % to disk (line 9), and mark them as clean (line 10).
    % If the flushing time does not exceed our time interval, the thread is put 
    % to sleep for the remaining time (lines 11-13). 
    % Then based on the state of the host at current simulated time, 
    % the algorithm continues or finishes the loop.
    Because periodical flushing is simulated as a background thread, it can happen concurrently
    with disk I/O initiated by the main thread. This parallelism is handled by the 
    storage model and reflected in simulated I/O time.

@@

    \begin{algorithm}\caption{File chunk read simulation of IOController}
    \label{alg:read}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{fs}{file size (assumed to fit in memory)}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
               \State disk\_read = min(cs, fs - mm.cached(fn))
               \State cache\_read = cs - disk\_read
            \State mm.flush(cs + disk\_read - mm.free\_mem - mm.evictable, fn) 
            \State mm.evict(cs + disk\_read - mm.free\_mem, fn) 
            \If {disk\_read $>$ 0}  \Comment{Read uncached data}    
                \State sm.read(disk\_read)  
                \State mm.add\_to\_cache(disk\_read, fn)     
            \EndIf
            \If {cache\_read $>$ 0} \Comment{Read cached}
                \State mm.cache\_read(cache\_read)  
            \EndIf
        \end{algorithmic}
    \end{algorithm}
    As mentioned previously, our model reads and writes file chunks in a
    round-robin fashion. To read a file chunk, simulated applications send
    chunk read requests to the IOController which processes them using
    Algorithm~\ref{alg:read}. First, we calculate the amount of uncached
    data that needs to be read from disk, and the remaining amount is read
    from cache (line 7-8). The amount of memory required to read the chunk
    is calculated, corresponding to a copy of the chunk in anonymous memory
    and a copy of the chunk in cache \tristan{Create a variable
    "required\_mem" in the algorithm to clarify}.
    \tristan{add a call to mm in the algorithm to increment anonymous memory. Mention in the experiments that anonymous memory is flushed after each task.}
    If there is not enough available memory, the Memory Manager is called
    to flush dirty data (line 9). If necessary, flushing is complemented by
    eviction (line 10). When called with negative arguments, functions
    flush and evict simply return and do not do anything. Then, if there is
    uncached data, a disk read is simulated with disk model (line 12)
    \tristan{this is not consistent with the new Figure 1 where disk reads
    are done by the mm}, and this data of the file is added to cache (line
    13).
    Finally, if cached data needs to be read, the Memory Manager is called
    to simulate a cache read  
    and update the corresponding data blocks accordingly (line 16).

    \begin{algorithm}\caption{File chunk write simulation of IOController}
    \label{alg:write}
        \small
        \begin{algorithmic}[1]
            \Input
                \Desc{cs}{chunk size}
                \Desc{fn}{file name}
                \Desc{mm}{MemoryManager object}
                \Desc{sm}{storage simulation model}
               \EndInput
            \State remain\_dirty = dirty\_ratio * mm.avail\_mem - mm.dirty
            \If {remain\_dirty $>$ 0} \Comment{Write with memory bandwidth}
                \State mm.evict(min(cs, remain\_dirty) - mm.free\_mem)
                \State mem\_amt = min(cs, mm.free\_mem)
                \State mm.write(fn, mem\_amt) 
            \EndIf
            \If {mem\_amt $<$ cs}  \Comment{Write with disk bandwidth}
                \State mm.flush(cs - mem\_amt)  
                \State mm.evict(cs - mem\_amt  - mm.free\_mem) 
                \State mm.add\_to\_cache(fn, min(cs - mem\_amt, mm.free\_mem)) \tristan{why is data written to cache with "addtocache" here and with "write" at line 10?}
                \State sm.write(cs - mem\_amt) \tristan{According to Fig 1 the IOController is not supposed to interact directly with sm}
            \EndIf

        \end{algorithmic}
    \end{algorithm}
    Algorithm~\ref{alg:write} describes our simulation of chunk writes in 
    the IOController. 
    % Data is written to cache until the amount of dirty data in the cache 
    % exceeds the dirty ratio times the amount of available memory. Data is then written to disk.
    Our algorithm initially checks the  amount of data that 
    can be written as dirty data given the dirty ratio (line 6).
    If this amount is greater than 0, the Memory Manager is requested to evict 
    data from cache if necessary (line 8) \tristan{the difference between available and free memory should be explained somewhere}.
    After eviction, the amount of data that can be written to 
    page cache is calculated (line 9), and a cache write is simulated (line 10).
    To write the remaining data, we flush and 
    evict data as possible so that this data can be 
    written to cache (line 13-14). 
    The Memory Manager does not guarantee that the amount of flushed 
    and evicted data is enough to write the remaining data. 
    In the case it is less than the required amount, the remaining data 
    being written to cache can be flushed and evicted right away. 
    Thus, the amount of remaining data written and kept in cache 
    after the write is limited by the amount free memory, and this amount 
    is added to cache (line 15) \tristan{the previous two sentences are very unclear, I don't understand them so I can't reformulate.}. 
    Finally, the disk write is simulated with the remaining data amount (line 16), 
    and the chunk write algorithm finishes \tristan{see comment in algorithm}.
